\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}
My first thanks go to my supervisors, Cristel Pelsser and Laurent Francis, for their guidance and support throughout this project. Cristel set up the UCLouvain server infrastructure and stayed available all year, both of which were essential to finishing this thesis. Laurent's expertise in sensor methodology and calibration kept the hardware work grounded in sound metrological practice.

Xavier Draye also helped me beyond just reading the final manuscript. As the professor for the LBIR1251 course, he helped me work out what were the requirements for this iteration of the project. So what to improve, what was already fine and which physiological metrics mattered. These conversations helped establish the scope of the project and kept it focused on the things that would make a difference for the students. I am grateful for his time and insights on this matter as well as for his reading of the final manuscript.

Thanks to Sebastien Toussaint for agreeing to sit on my jury and for taking the time to review this work.

Tom Klauner at the WELCOME lab helped me through the sensor calibration sessions and was generous with his time throughout that process, for which I am very grateful.

Alexandre and Arnaud Wery helped me in designing the 3D-printed casing and soldering the station's components. I therefore want to thank them both. 

Finally, I want to thank my family: my mother Virginie Goffaux, my father Eric Montariol, and Colin Larcombe, for their careful proofreading and for their support throughout this year.

