\chapter{Use of artificial intelligence tools}
\label{ch:appendix-ai}

In accordance with UCLouvain's policy on the use of generative artificial intelligence and academic transparency standards, the following AI tools were used during the development of this Master's thesis for the following purposes:
\begin{itemize}
    \item \textbf{Code assistance and debugging - Claude Code:} Was used as a pair-programming assistant throughout development to accelerate the writing of acquisition scripts and boilerplate code, to help debug I$^2$C driver issues and to help propose fixes to implementation problems. All of the suggested code was reviewed and tested on the physical hardware and was manually validated before integration into the codebase.
    \item \textbf{Mathematical assistance - Claude:} Was used to verify the calibration offset calculations and check various mathematical formulas. All results were independently and manually checked.
    \item \textbf{LaTeX formatting - Claude Code:} Was used to generate TikZ diagram scaffolding and tabular structure templates as well as general formatting assistance. All generated markup was reviewed, fact checked and adapted to match the thesis content.
    \item \textbf{Writing and proofreading - Claude:} Was used to correct grammar and spelling errors and to improve general clarity. All of the text in this manuscript reflects my own analysis, methodology and conclusions. There are no AI-generated passages that appear word-for-word in the final manuscript that weren't reviewed and revised.
\end{itemize}
All AI-assisted contributions were reviewed, tested where applicable and validated to ensure technical accuracy and scientific integrity.
